---
title: "Workshop #2: Construct the alternatives"
format: html
---

```{r slides, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
```

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("../slides/workshop02")
```

<div class="container" style="position: relative;overflow: hidden;width: 100%;min-height:600px;">
<iframe src="../slides/workshop02.html" style="position=absolute;top: 0;left: 0;bottom: 0;right: 0;width: 100%;min-height:600px;"></iframe>
</div>

# Introduction

> The following text is based on the assigned readings for this week. Refer to the syllabus for the detailed information.

Welcome to the our second workshop! This handout covers **Steps 2, 3, and 4** from Eugene Bardach and Eric M. Patashnik’s _A Practical Guide for Policy Analysis_ (2020). These steps are crucial for assembling evidence, constructing policy alternatives, and selecting evaluation criteria for your policy analysis paper.

# Step 2: Assemble Some Evidence

## Importance of Evidence in Policy Analysis

All time spent on policy analysis revolves around two activities: **thinking** and **hustling data** that can be transformed into **evidence**. Effective evidence collection is essential, especially given the real-world constraints like time pressure, which can jeopardize the quality of your analysis.

> *"Evidence is information that affects the existing beliefs of important people (including yourself) about significant features of the problem you are studying and how it might be solved or mitigated."*  
> — Bardach & Patashnik, 2020, p. 36

## Purposes of Assembling Evidence

1. **Assessing the Problem:** Understand the nature and extent of the problem you're addressing.
2. **Understanding the Policy Context:** Gain insights into specific features of the policy situation, such as agency workloads, budget figures, political ideologies, and demographic changes.
3. **Evaluating Existing Policies:** Examine policies that have worked in similar situations elsewhere.

### Example

Suppose you're analyzing the issue of smoking prevalence in different countries. Differential smoking rates can provide evidence about varying levels of concern for personal health, informing potential policy interventions.

## Efficient Data Collection

- **Focus on Relevance:** Continuously ask yourself what information you need and why.
- **Avoid Irrelevant Data:** Collecting data that cannot be developed into meaningful evidence is a common mistake.
- **Value vs. Cost:** Weigh the cost of obtaining evidence against its potential value in improving policy decisions.

### Tips for Data Collection

- **Start Early:** Coordinating with busy stakeholders requires time.
- **Review Available Literature:** Utilize online resources to access existing research.
- **Use Analogies and Best Practices:** Look for solutions that worked in different contexts or consider unconventional approaches.

## Building Credibility and Consensus

Engage with potential critics to incorporate diverse perspectives, ensuring that your analysis is balanced and comprehensive.

# Step 3: Construct the Alternatives

## Defining Alternatives

**Alternatives** refer to different policy options or courses of action aimed at solving or mitigating the problem. These can range from single actions to combinations of strategies.

> *"Policy analysts use the term alternative ambiguously: sometimes it means one choice that implies forgoing another, and sometimes it means simply one more policy action that might help to solve or mitigate a problem, perhaps in conjunction with other alternatives."*  
> — Bardach & Patashnik, 2020, p. 42

## Comprehensive vs. Focused Alternatives

- **Start Comprehensive:** Initially, list a wide range of potential alternatives.
- **Narrow Down:** Gradually eliminate less viable options to focus on the most promising ones.

### Example

If addressing heroin use reduction, you might consider:
- **Methadone Maintenance:** Expanding access to treatment programs.
- **Law Enforcement Pressure:** Increasing policing efforts to deter use.
- **Drug Education:** Implementing widespread educational campaigns.

## Sources for Generating Alternatives

1. **Proposals from Political Actors:** Identify what key stakeholders are already considering.
2. **Generic Strategies:** Refer to checklists like Appendix A, “Things Governments Do,” to inspire innovative solutions.
3. **Unconventional Solutions:** Encourage creative thinking by considering out-of-the-box ideas.

### Creative Techniques

- **"If cost were no object":** Imagine ideal solutions without budget constraints.
- **Analogous Contexts:** Apply successful strategies from different fields or regions.
- **Challenge Assumptions:** Regularly ask “why not” to explore new possibilities.

## Refining Alternatives

- **Behavioral Detail:** Clearly define what actions each alternative entails.
- **Multistage Analysis:** Consider policies as part of a dynamic, ongoing process rather than one-off decisions.

# Step 4: Select the Criteria

## Understanding Criteria

Criteria are the standards used to evaluate the potential outcomes of each alternative. They bridge the gap between the analytic and evaluative aspects of policy analysis.

> *"The most important step for introducing values and philosophy into the policy analysis, because some possible 'criteria' are evaluative standards used to judge the goodness of the projected policy outcomes that are associated with each of the alternatives."*  
> — Bardach & Patashnik, 2020, p. 53

## Types of Criteria

1. **Evaluative Criteria:** Standards used to assess the desirability of outcomes (e.g., efficiency, equity).
2. **Practical Criteria:** Practical considerations related to policy implementation (e.g., legality, political acceptability).

### Common Evaluative Criteria

- **Efficiency:** Maximizing aggregate welfare or public interest.
- **Effectiveness:** The degree to which the policy achieves its objectives.
- **Equity:** Fair distribution of benefits and burdens.
- **Political Acceptability:** The extent to which the policy is supported by stakeholders.

### Example

For a policy aimed at reducing homelessness:
- **Principal Criterion:** Minimize the number of homeless families.
- **Secondary Criteria:** Cost-effectiveness, political feasibility, and social equity.

## Selecting and Defining Criteria

1. **Primary Criterion:** Directly related to solving the core problem.
2. **Secondary Criteria:** Additional factors that influence the desirability of outcomes.

### Steps to Define Criteria

- **Maximize/Minimize:** Clearly state whether you aim to increase or decrease a particular value.
- **Operationalize:** Define metrics for each criterion to ensure clarity and measurability.
  
  | **Criterion** | **Metric** |
  |---------------|------------|
  | Efficiency    | Cost per homeless family reduced |
  | Effectiveness | Percentage decrease in homelessness |
  | Equity        | Distribution of benefits across demographics |

## Avoiding Common Mistakes

- **Do Not Confuse Alternatives with Criteria:** Alternatives are actions; criteria are standards for evaluation.
  
  > *"Alternatives are courses of action, whereas criteria are mental standards for evaluating the results of action."*  
  > — Bardach & Patashnik, 2020, p. 69

## Weighting Criteria

- **Relative Importance:** Determine the weight of each criterion based on their importance to the policy goals.
- **Balancing Values:** Address conflicts between criteria by prioritizing based on overarching values or philosophical considerations.

### Approaches to Weighting

1. **Political Process:** Let existing political frameworks determine weights.
2. **Analyst Imposition:** Adjust weights based on fairness and underrepresented interests.

## Practical Application

When selecting criteria for your analysis:
- **Group Positive and Negative Criteria Separately:** For clarity in evaluation.
- **Specify Metrics:** Ensure each criterion has a clear, measurable indicator.
- **Focus on Clarity and Relevance:** Criteria should be directly related to the policy objectives and practical considerations.

# Conclusion

By meticulously assembling evidence, constructing a comprehensive set of alternatives, and thoughtfully selecting evaluation criteria, you lay a strong foundation for your policy analysis. Use this handout as a guide to navigate these critical steps effectively in your project analysis paper.


