---
title: "Workshop #6: Individual Memo"
format: html
---

```{r slides, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
```

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("../slides/workshop06")
```

<div class="container" style="position: relative; overflow: hidden; width: 100%; min-height:600px;">
  <iframe src="../slides/workshop06.html" style="position:absolute; top: 0; left: 0; bottom: 0; right: 0; width: 100%; height:100%;"></iframe>
</div>

## Workshop Overview

In this final workshop, we focus on crafting Individual Memos, an essential skill for effective policy analysis. The memo provides actionable recommendations, supported by analysis and evidence, in a concise and clear format tailored to the needs of your client.

## Objectives

- Analyze and critique sample memos to identify strengths and areas for improvement.
- Develop concise writing strategies to create impactful, focused memos.
- Learn tools for structuring arguments, addressing counterarguments, and aligning recommendations with client priorities.

## Activities

- Memo Analysis: Review and critique sample memos to understand best practices.
- Interactive Exercise: Roleplay as an analyst, summarizing findings and making recommendations.
- Grandma Bessie Test: Craft and refine a 1-minute elevator pitch for your memo.

## Key Takeaways

- Clarity is crucial: Place the recommendation upfront, use simple language, and structure the memo for readability.
- Brevity matters: Stay within the word count by prioritizing essential ideas.
- Anticipate client needs: Align every argument with what matters most to the client.
- Use visual aids: Simplify complex arguments and comparisons with tables, graphs, or matrices.
- Practice storytelling: Frame the memo as a narrative—problem, options, recommendation, and evidence.

Prepare to leave this workshop equipped with tools to create effective, polished policy memos that make a lasting impact!

